# Reproducible Research: Peer Assessment 1

## Setting ECHO=TRUE Globally

```{r setoptions, echo=TRUE}
opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data

```{r setting working directory}
getwd()
```
```{r loading data}
data <- read.csv("activity.csv")
```


### Make a Histogram of the total number of steps taken each day

```{r histogram of sums of steps}
sumOfSteps <- aggregate(steps~date, data, sum)
hist(sumOfSteps$steps,
     breaks = 25,
     col = "red",
     xlab = "Total steps per day",
     main = "Histogram of Steps per Day")
```

### What is the mean and median of the total number of steps taken per day?

```{r mean and median of steps}
mean(sumOfSteps$steps)
median(sumOfSteps$steps)
```

## What is the average daily activity pattern?

### Time series plot of 5-minute interval vs. average steps

```{r Average daily activity}
steps5Min <- aggregate(steps~interval, data, mean)
plot(steps5Min$interval, steps5Min$steps, type="l",
     main = "Average Daily Activity",
     xlab = "5 minutes intervals",
     ylab = "Average steps per day")
```

### Which 5 minute interval on average contains the maximum number of steps?

```{r maximum number of steps per interval}
max(steps5Min$steps)
steps5Min[steps5Min$steps > 206,]
```

## Imputing missing values

### Calculate and report total number of missing values in the dataset

```{r number of rows with missing values}
sum(complete.cases(data)) ## cases without NA in any column
sum(!is.na(data$steps)) ## rows without missing values in column steps
sum(is.na(data$steps))  ## rows with missing values in column steps
sum(is.na(data$date))  ## rows with missing values in column date
sum(is.na(data$interval))  ## rows with missing values in column interval
```
### There are 2304 cases with missing values, all of them in column steps.

### Filling Missing Values with imputation

```{r imputation of Missing Values}

#library(Amelia) ## loads program for imputation
#AmeliaView() ## loads GUI to do the imputation

## calculates means of steps in the whole file to be used as maximum value in the imputation 
meanOfSteps <- mean(data$steps, na.rm=TRUE) ## the mean of Steps is 37.38

## loading new dataset with imputed Missing Values
data2 <- read.csv("activity-imp1.csv")
```
### Make Histogram of the total step taken in the day and calculate mean and median

```{r new histogram of sums of steps}
sumOfSteps2 <- aggregate(steps~date, data2, sum)
hist(sumOfSteps2$steps,
     breaks = 25,
     col = "blue",
     xlab = "Total steps per day",
     main = "New Histogram of Steps per Day")
```

### What is the mean and median of the total number of steps taken per day?

```{r new mean and median of steps}
mean(sumOfSteps2$steps)
median(sumOfSteps2$steps)
```
### Imputing the Missing data did not modify the histogram, mean and median significantly

## Are there differences in activity patterns between weekdays and weekends?

### Adding the column days and converting dates to factors (weekday - weekend)

```{r weekday - weekend factors}
library(lubridate)

data2 <- read.csv("activity-imp1.csv")

data2["days"] <- (data2date <- as.Date(data2$date))
data2["days"] <- wday(data2$days, label=TRUE)

days <- c("Sun", "Mon", "Tues", "Wed", "Thurs", "Fri", "Sat")
#days <- as.factor(days)


data2$period <- NA

data2$period[data2$days=="Sat"] <- as.character("Weekend") 
data2$period[data2$days=="Sun"] <- as.character("Weekend") 
data2$period[data2$days=="Mon"] <- as.character("Weekday") 
data2$period[data2$days=="Tues"] <- as.character("Weekday")  
data2$period[data2$days=="Wed"] <- as.character("Weekday")  
data2$period[data2$days=="Thurs"] <- as.character("Weekday")  
data2$period[data2$days=="Fri"] <- as.character("Weekday")  

data2$period2 <- (data2$period)



```
## Make panel plot with time series of 5 minute-interval (x) and the average number of steps (y)
```{r Weekday/Weekend Average daily activity}
steps5Min <- aggregate(steps~interval, data2, FUN=mean)

library(lattice)
xyplot(steps5Min$steps~steps5Min$interval|(data2$period2) , type="l",
     layout=c(1,2),
     main = "Weekday/Weekend Activity",
     xlab = "Interval",
     ylab = "Number of steps")
```
